---
title: "penguin_analysis"
output: html_document
date: "2024-10-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(palmerpenguins)
library(janitor)
library(here)
```

```{r}
here::here()
#checking working directory
#less buggy than getwd()

head(penguins_raw)
```

```{r}
colnames(penguins_raw)

```
There are lots of issues with column names - you cannot just edit the names in excel
It is important to preserve raw data as a csv file 

Going to use the here library to help us with the file path. We are telling it to use the named folder
and the filename penguins_raw.csv

```{r}

write.csv(penguins_raw, here("data", "penguins_raw.csv"))

```

Now removing columns:

```{r}
colnames(penguins_raw)

library(dplyr)
penguins_raw <- select(penguins_raw, -Comments)

colnames(penguins_raw)

```
Can do the same for two columns at once:

```{r}

#THIS IS BAD CODE 
#YOU HAVE JUST OVERWRITTEN THE OBJECT
colnames(penguins_raw)

penguins_raw <- select(penguins_raw, -Comments)
penguins_raw <- select(penguins_raw, -starts_with("Delta"))

colnames(penguins_raw)


```

So we need to load back in our safe copy of penguins code...

```{r}
penguins_raw <- read.csv(here("data", "penguins_raw.csv"))

#checking

colnames(penguins_raw)

#make new variable, remove comments

penguins_clean <- select(penguins_raw, -Comments)

#removing delta columns - STILL BAD CODE
#same issue of overwriting - we have already removed columns above 
penguins_clean <- select(penguins_clean, -starts_with("Delta"))

```

Using piping to remove comments and delta columns

```{r}

library(tidyverse)
library(janitor)

penguins_clean <- penguins_raw %>%
  select(-"Comments") %>%
  select(-starts_with("Delta")) %>% #need to make sure steps are in the right order
  clean_names()

colnames(penguins_clean)
#cleans column names

```

All code together:

```{r}

library(tidyverse)
library(janitor)
library(palmerpenguins)
library(here)

#writing a safe copy:

#checking column names:

#loading data from csv

#cleaning it up using the pipe



```

Now going to make a function

```{r}

#called cleaning_penguin columns

cleaning_penguin_columns <- function(raw_data) { #raw_data could have a different name
  raw_data %>%
    clean_names %>% #janitor function
    remove_empty(c("rows", "cols")) %>% #c joins them together
    select(-comments) %>% #might have to remove to get it to work
    select(-starts_with("delta")) %>% #lowercase as the data will already have been cleaned
    print("Removed empty columns and rows, cleaned column names, removed comments and delta columns")
}

#loading the data again

penguins_raw <- read.csv(here("data", "penguins_raw.csv"))

#check the column names
colnames(penguins_raw)

#running the function
library(dplyr)
penguins_clean <- cleaning_penguin_columns(penguins_raw)

#checking the output
colnames(penguins_clean)

#in reality the function would be more applicable in a different setting 
```
Now we have clean data we now need to save it

```{r}
write.csv(penguins_clean, here("data", "penguins_clean.csv"))

```


Can also move the cleaning function to a separate script
Going to make a new subfolder called functions where we can put our reusable co

#in the console:
file.create(here"functions", "cleaning.R")

then went onto canvas and downloaded the cleaning.R function file
need the R markdown file

```{r}

source(here("functions", "cleaning.R"))

#loading functions from a different file

```


Using function 
